Estimation of flow trajectories in a multi-lines network

### 2023-02-08 latex_files/applied_network_science_article/article_v0.tex
commit ****
rewriting method

Let the transportation flow $n_{st}$ denote the number of passengers following an $st$-trip, that is entering the network at $s$ and leaving the network at $t$ by using the shortest path.

*** Also, 
at the junction level, let $\tilde{n}_{ST}$ denote the number of passengers following an $ST$-trip. As implied by the end of section (\ref{Line edges, transfer edges and trips}), 
\begin{equation}
\label{equivnntilde}
\tilde{n}_{ST}=\sum_{s\in S}\sum_{t\in T}n_{st} \qquad\qquad\qquad n_{st}=\sum_{ST}\tilde{n}_{ST}\: \delta_{s,\sigma(S,T)}\: \delta_{t,\tau(S,T)}
\end{equation}
Definitions (\ref{edgetrip}) and (\ref{edgetripjunction}) yield
\begin{equation}
\label{equationGG}
x_{ij}=\sum_{st}\chi_{ij}^{st}\:  n_{st}=\sum_{ST}\tilde{\chi}_{ij}^{ST}\:  \tilde{n}_{ST}
\end{equation}
Among the passengers embarking in $i$, some transfer from another line, and some others enter into the network: 
\begin{equation}
\label{entrer}
a_i=z_{\bullet i}+n_{i\bullet}
\end{equation}
where  ``$\bullet$" denotes the summation over the replaced index, as in $n_{i\bullet}=\sum_{j=1}^l n_{ij}$. Similarly, among the passengers disembarking in $i$, some transfer to another line, and some others leave the network: 
\begin{equation}
\label{sortir}
b_i=z_{i\bullet}+n_{\bullet i}
\end{equation}
By construction
\begin{displaymath}
a_{\bullet}=b_{\bullet}=z_{\bullet\bullet}+n_{\bullet\bullet}
\end{displaymath}
where $n_{\bullet\bullet}$ counts the number of passengers, and $z_{\bullet\bullet}$ counts the number of transfers. $z_{\bullet\bullet}/n_{\bullet\bullet}$  is the average number of transfers per passenger.

@@

As explained in section \ref{Lines and junctions}, transfers can only occur at junctions, that is $z_{ij}>0$ implies $S(i)=S(j)$. As $z_{ii}=0$ (no traveller is supposed to disembark and re-embark later at the same stop), and $z_{ii\text{-}}=0$ (where $i\text{-}$ is the stop opposite to $i$, when existing), transfers can occur at true junctions only.

At the junction level, let 
\begin{displaymath}
\tilde{a}_I=\sum_{i\in I} a_i\enspace,\enspace\tilde{b}_I=\sum_{i\in I} b_i\enspace,\enspace\tilde{z}_{\bullet I}=\sum_{i\in I} z_{\bullet i}\enspace,\enspace\tilde{z}_{I\bullet}=\sum_{i\in I} z_{i\bullet}\enspace,\enspace\tilde{n}_{I\bullet}=\sum_{i\in I} n_{i\bullet}\enspace,\enspace\tilde{n}_{\bullet I}=\sum_{i\in I} n_{\bullet i}
\end{displaymath}
 respectively denote the number of embarking (disembarking)  passengers at junction~$I$; of embarking passengers after a transfer and of disembarking passengers before a transfer at  $I$; of passengers starting (ending) their trip at $I$. Then 
\begin{equation}
\label{junctionLevel}
\tilde{a}_I=\tilde{z}_{\bullet I}+\tilde{n}_{I\bullet}\qquad\qquad \tilde{b}_I=\tilde{z}_{I\bullet}+\tilde{n}_{\bullet I}\qquad\qquad \tilde{z}_{\bullet I}=\tilde{z}_{I\bullet}\le \min(\tilde{a}_I,\tilde{b}_I)\enspace. 
\end{equation}

\subsection{Statement of the problem}
Automatic passenger counters measure the number of passengers entering and leaving buses at each stop [Boyle, 1998], that is $\mathbf{a}$ and $\mathbf{b}$. Also, the geometry of the network permits to derive the edge-trip incidence matrix $\bm{\chi}$ (\ref{edgetrip}), as well as the partition $\mathcal{S}$  of stops nested within junctions.

Intra-line edge flows $\mathbf{Y}=(y_{ij})$ can be determined by (\ref{bilan1ligne}), but transfer edge flows $\mathbf{Z}=(z_{ij})$ are, here and typically, unknown. The objective is to estimate the $l\times l$ transportation flow $\mathbf{N}=(n_{st})$ at the stop level, or equivalently  the $L\times L$  transportation flow  $\tilde{\mathbf{N}}=(\tilde{n}_{ST})$ at the junction level. Many consistent solutions coexist in general, even for a single line with no transferts (section \ref{Single line}).

This issue of incompletely observed data can be tackled by the maximum entropy formalism: let $f_{st}=n_{st}/n_{\bullet\bullet}$ be the proportion of $st$-trips (empirical distribution) and let $g_{st}$ be some prior guess on its shape (theoretical distribution). Let $\mathcal{F}$ consist of the empirical distributions $\mathbf{f}$ satisfying $R$ linear constraints of the form $\sum_{st}f_{st}\, o^r_{st}=\bar{o}^r$ for $r=1,\ldots, R$. Then the minimum Kullback-Leibler divergence 
\begin{displaymath}
\min_{\mathbf{f}\in\mathcal{F}}\;  \sum_{st}f_{st}\log \frac{f_{st}}{g_{st}}
\end{displaymath}
is reached for the maximum entropy solution  
\begin{equation}
\label{MaxEnt}
f^0_{st}=\frac{1}{Z(\bm{\lambda})}\: g_{st}\exp(-\sum_{r=1}^R \lambda_r o^r_{st})\qquad\qquad Z(\bm{\lambda})=\sum_{st}g_{st}\exp(-\sum_{r=1}^R \lambda_r o^r_{st})
\end{equation}
where the Lagrange multipliers $\bm{\lambda}$ are determined  so as to satisfy the $R$ linear constraints (**see e.g. Bavaud and *** and references therein).

\subsection{Statement of the problem (version 2, added by GG)}
Let $f_{st}=n_{st}/n_{\bullet\bullet}$ be the proportion of $st$-trips (empirical distribution) and let $g_{st}$ be some prior guess on its shape (theoretical distribution). Also suppose a prior guess of row and margin constraints for $f_{st}$, respectively $\alpha_s$ and $\beta_t$, which has to verify 
\begin{equation}
	\label{init_margin}
	\frac{\alpha_i}{\alpha_j} = \frac{a_i}{a_j} \qquad \frac{\beta_i}{\beta_j} = \frac{b_i}{b_j}  \qquad \forall i, j \text{ which are not true junctions.} \\
\end{equation}
Setting these gives us
\begin{equation}
	\label{flow_sum}
	n_{\bullet \bullet} = \frac{a_i}{\alpha_i} = \frac{b_i}{\beta_i} \qquad \forall i \text{ which are not true junctions.}  
\end{equation}
We are interested in finding
\begin{align}
	\label{constr_MaxEnt}
	\min_{\mathbf{f}\in\mathcal{F}} &\; \sum_{st}f_{st}\log \frac{f_{st}}{g_{st}}, \notag \\
	s.t. &\; \sum_t f_{st} = \alpha_s, \notag \\
	&\; \sum_s f_{st} = \beta_t.
\end{align}
The Lagragian is
\begin{equation}
	L = \sum_{st}f_{st}\log \frac{f_{st}}{g_{st}} - \sum_s \lambda_s (\alpha_s - \sum_t f_{st}) - \sum_t \mu_t (\beta_t - \sum_s f_{st}), \notag
\end{equation}
which gives, after deriving and setting to zero,
\begin{equation}
	\label{Sol}
	f_{st} = \phi_s \psi_t g_{st} \qquad \text{with } \phi_s := \exp(- 1 - \lambda_s) \text{, } \psi_t := \exp(- \mu_t).
\end{equation}
Using constraints in (\ref{constr_MaxEnt}), we find
\begin{equation}
	\label{Sol_LagMult}
	\phi_s = \frac{\alpha_s}{\sum_t \psi_t g_{st}}, \qquad \psi_t = \frac{\beta_t}{\sum_s \phi_s g_{st}}, 
\end{equation}
which gives an \emph{iterative fitting algorithm}. Starting with any $\psi^{(0)}_t > 0$, we do
\begin{equation}
	\label{Iterative fitting}
	\phi^{(\iota)}_s = \frac{\alpha_s}{\sum_t \psi^{(\iota)}_t g_{st}}, \qquad \psi^{(\iota + 1)}_t = \frac{\beta_t}{\sum_s \phi^{(\iota)}_s g_{st}}, 
\end{equation}

@@

\subsection{Margin distributions update (added by GG)}

@@

\subsection{Prior distribution update (added by GG)}

Transfer edges $(i, j)$ where $z_{i \bullet} > (1 - \theta)b_i$ or $z_{\bullet j} > (1 - \theta)a_j$ typically denote an overflowing tansfer flow regarding the $g_{st}$ prior distribution.  The latter must be adjusted in order to have $\mathbf{Z}$ comptible with constraints. For any edge $(i, j)$, let us compute the \emph{flow ratio} $r_{ij}$, with 
\begin{equation}
	\label{flow_ratio}
	r_{ij} = \max \left(1, \frac{z_{i \bullet}}{(1 - \theta)b_i}, \frac{z_{\bullet j}}{(1 - \theta)a_j} \right),
\end{equation}
where $r_{ij} > 1$ typically denotes an overflow through edge $(i, j)$. This flow ratio can be traced back to origin-destination couples $s, t$ in order to compute the \emph{orgin-destination flow ratio} $\bar{r}_{st}$ 
\begin{equation}
	\label{st_flow_ratio}
	\bar{r}_{st} = \max_{ij} \chi_{ij}^{st} r_{ij},
\end{equation}
where again $\bar{r}_{st} > 1$ denotes an overflow between $s$ and $t$. To ajdust the flow, we can devide the previous flow by this ratio
\begin{equation}
	\label{update_flow}
	\widetilde{n}_{st} =\frac{n_{st}}{\bar{r}_{st}}
\end{equation}
And the new prior distribution should follow 
\begin{equation}
	\label{update_distrib}
	\widetilde{g}_{st} = \frac{\left( \frac{\widetilde{n}_{st}}{\phi_s \psi_t} \right)}{\sum_{s',t'} \left( \frac{\widetilde{n}_{s',t'}}{\phi_{s'} \psi_{t'}} \right)}
\end{equation}

With the new prior distribution $\widetilde{g}_{st}$ and the new margin distributions $\widetilde{\alpha}_s$, $\widetilde{\beta}_t$, we can begin a new iteration step by going through 2.5 -2.7 again.

>>>>

Let the transportation flow $n_{st}$ denote the number of passengers following an $st$-trip, that is entering the network at $s$ and leaving the network at $t$ by using the shortest path. One gets from (\ref{edgetrip}) 
\begin{equation}
\label{equationGG}
x_{ij}=\sum_{st}\chi_{ij}^{st}\:  n_{st}
\end{equation}
Among the passengers embarking in $i$, some transfer from another line, and some others enter into the network: 
\begin{equation}
\label{entrer}
a_i=z_{\bullet i}+n_{i\bullet}
\end{equation}
where  ``$\bullet$" denotes the summation over the replaced index, as in $n_{i\bullet}=\sum_{j=1}^l n_{ij}$. Similarly, among the passengers disembarking in $i$, some transfer to another line, and some others leave the network: 
\begin{equation}
\label{sortir}
b_i=z_{i\bullet}+n_{\bullet i}
\end{equation}
By construction
\begin{displaymath}
a_{\bullet}=b_{\bullet}=z_{\bullet\bullet}+n_{\bullet\bullet}
\end{displaymath}
where $n_{\bullet\bullet}$ counts the number of passengers, and $z_{\bullet\bullet}$ counts the number of transfers. $z_{\bullet\bullet}/n_{\bullet\bullet}$  is the average number of transfers per passenger.

@@

As explained in section \ref{Lines and junctions}, transfers can only occur at junctions, that is $z_{ij}>0$ implies $j\in S_i$. In particular,  $z_{ii}=0$ : no traveller is supposed to disembark and re-embark later at the same stop.

\subsection{Statement of the problem and solution method}
Automatic passenger counters measure the number of passengers entering and leaving buses at each stop [Boyle, 1998], that is $\mathbf{a}$ and $\mathbf{b}$. Also, the geometry of the network permits to derive the edge-trip incidence matrix $\bm{\chi}$ defined in (\ref{edgetrip}).

Intra-line edge flows $\mathbf{Y}=(y_{ij})$ can be determined by (\ref{bilan1ligne}), but transfer edge flows $\mathbf{Z}=(z_{ij})$ are, here and typically, unknown. The objective is to estimate the $l\times l$ transportation flow $\mathbf{N}=(n_{st})$. Many consistent solutions coexist in general, even for a single line with no transferts (section \ref{Single line}). This issue of incompletely observed data can be tackled by the maximum entropy formalism (*** ref***).

Let $f_{st}=n_{st}/n_{\bullet\bullet}$ be the proportion of $st$-trips (empirical distribution) and let $g_{st}$ be some prior guess on its shape (theoretical distribution). 
Assuming some reasonable initial prior $g_{st}$, 
\begin{itemize}
\item[(1)] we shall first suppose that the empirical margins $\alpha_s=f_{s\bullet}$ and $\beta_t=f_{\bullet t}$ are known.  
Then $f_{st}$ can be determined as the maximum entropy solution (section \ref{maxenso}), i.e. as the distribution closest to $g_{st}$ in the Kullback-Leibler divergence sense under the margin constraints
\item[(2)] then (section \ref{marginup}), the margins will be updated to $\tilde{\alpha}_s$ and $\tilde{\beta}_t$   by requiring a minimum proportion $\theta\in (0,1)$ of passengers entering/leaving the network at each stop, as well as avoiding transfer overflow exceeding the embarking and disembarking counts at each stop
 \item[(3)] finally (section \ref{priorup}), the prior will be updated to $\tilde{g}_{st}$ by contracting, if necessary, the priors $g_{st}$ associated to overflows. 
\end{itemize}
With the new prior distribution $\widetilde{g}_{st}$ and the new margin distributions $\widetilde{\alpha}_s$, $\widetilde{\beta}_t$, we can iterate the 
 the above steps, until convergence. The only free parameter is $\theta$, which will be discussed in section *** . This solution method is reminiscent, but quite distinct of the ME algorithm ...***

\subsubsection{***Maximum entropy solution} 
Let $\mathcal{F}$ consist of the empirical distributions $\mathbf{f}$ satisfying $R$ linear constraints of the form $\sum_{st}f_{st}\, o^r_{st}=\bar{o}^r$ for $r=1,\ldots, R$. Then the minimum Kullback-Leibler divergence 
\begin{displaymath}
\min_{\mathbf{f}\in\mathcal{F}}\;  \sum_{st}f_{st}\log \frac{f_{st}}{g_{st}}
\end{displaymath}
is reached for the maximum entropy solution  
\begin{equation}
\label{MaxEnt}
f^0_{st}=\frac{1}{Z(\bm{\lambda})}\: g_{st}\exp(-\sum_{r=1}^R \lambda_r o^r_{st})\qquad\qquad Z(\bm{\lambda})=\sum_{st}g_{st}\exp(-\sum_{r=1}^R \lambda_r o^r_{st})
\end{equation}
where the Lagrange multipliers $\bm{\lambda}$ are determined  so as to satisfy the $R$ linear constraints (**see e.g. Bavaud and *** and references therein).

\subsubsection{Maximum entropy solution ***Statement of the problem (version 2, added by GG)}
\label{maxenso}
Let $f_{st}=n_{st}/n_{\bullet\bullet}$ be the proportion of $st$-trips (empirical distribution) and let $g_{st}$ be some prior guess on its shape (theoretical distribution). 
Also suppose a prior guess of row and margin constraints for $f_{st}$, respectively $\alpha_s$ and $\beta_t$, which has to verify 
\begin{equation}
	\label{init_margin}
	\frac{\alpha_i}{\alpha_j} = \frac{a_i}{a_j} \qquad \frac{\beta_i}{\beta_j} = \frac{b_i}{b_j}  \qquad \forall i, j \text{ which are not true junctions.} \\
\end{equation}
Setting these gives us
\begin{equation}
	\label{flow_sum}
	n_{\bullet \bullet} = \frac{a_i}{\alpha_i} = \frac{b_i}{\beta_i} \qquad \forall i \text{ which are not true junctions.}  
\end{equation}
We are interested in finding
\begin{align}
	\label{constr_MaxEnt}
	\min_{\mathbf{f}\in\mathcal{F}} &\; \sum_{st}f_{st}\log \frac{f_{st}}{g_{st}}, \notag \\
	s.t. &\; \sum_t f_{st} = \alpha_s, \notag \\
	&\; \sum_s f_{st} = \beta_t.
\end{align}
The Lagragian is
\begin{equation}
	L = \sum_{st}f_{st}\log \frac{f_{st}}{g_{st}} - \sum_s \lambda_s (\alpha_s - \sum_t f_{st}) - \sum_t \mu_t (\beta_t - \sum_s f_{st}), \notag
\end{equation}
which gives, after deriving and setting to zero,
\begin{equation}
	\label{Sol}
	f_{st} = \phi_s \psi_t g_{st} \qquad \text{with } \phi_s := \exp(- 1 - \lambda_s) \text{, } \psi_t := \exp(- \mu_t).
\end{equation}
Using constraints in (\ref{constr_MaxEnt}), we find
\begin{equation}
	\label{Sol_LagMult}
	\phi_s = \frac{\alpha_s}{\sum_t \psi_t g_{st}}, \qquad \psi_t = \frac{\beta_t}{\sum_s \phi_s g_{st}}, 
\end{equation}
which gives an \emph{iterative fitting algorithm}. Starting with any $\psi^{(0)}_t > 0$, we do
\begin{equation}
	\label{Iterative fitting}
	\phi^{(\iota)}_s = \frac{\alpha_s}{\sum_t \psi^{(\iota)}_t g_{st}}, \qquad \psi^{(\iota + 1)}_t = \frac{\beta_t}{\sum_s \phi^{(\iota)}_s g_{st}}, 
\end{equation}

@@

\subsubsection{Margin distributions update}
\label{marginup}

@@

\subsubsection{Prior distribution update (added by GG)}
\label{priorup}
Transfer edges $(i, j)$ where $z_{i \bullet} > (1 - \theta)b_i$ or $z_{\bullet j} > (1 - \theta)a_j$ typically denote an overflowing tansfer flow regarding the $g_{st}$ prior distribution.  The latter must be adjusted in order to have $\mathbf{Z}$ comptible with constraints. For any edge $(i, j)$, let us compute the \emph{flow ratio} $r_{ij}$, with 
\begin{equation}
	\label{flow_ratio}
	r_{ij} = \max \left(1, \frac{z_{i \bullet}}{(1 - \theta)b_i}, \frac{z_{\bullet j}}{(1 - \theta)a_j} \right),
\end{equation}
where $r_{ij} > 1$ typically denotes an overflow through edge $(i, j)$. This flow ratio can be traced back to origin-destination couples $s, t$ in order to compute the \emph{orgin-destination flow ratio} $\bar{r}_{st}$ 
\begin{equation}
	\label{st_flow_ratio}
	\bar{r}_{st} = \max_{ij} \chi_{ij}^{st} r_{ij},
\end{equation}
where again $\bar{r}_{st} > 1$ denotes an overflow between $s$ and $t$. To ajdust the flow, we can devide the previous flow by this ratio
\begin{equation}
	\label{update_flow}
	\widetilde{n}_{st} =\frac{n_{st}}{\bar{r}_{st}}
\end{equation}
And the new prior distribution should follow 
\begin{equation}
	\label{update_distrib}
	\widetilde{g}_{st} = \frac{\left( \frac{\widetilde{n}_{st}}{\phi_s \psi_t} \right)}{\sum_{s',t'} \left( \frac{\widetilde{n}_{s',t'}}{\phi_{s'} \psi_{t'}} \right)}
\end{equation}
